\documentclass{article}
\usepackage{hyperref,relsize,,epsfig,a4,amsmath,amssymb}
\usepackage[latin1]{inputenc}
% required by ptex2tex:
\usepackage{graphicx,hyperref,relsize,fancyvrb,epsfig}
\usepackage{a4,amsmath,amssymb,framed,subfigure}
\usepackage[usenames]{color}
\begin{document}
\title{Doconce: Document Once, Include Everywhere}

\author{Hans Petter Langtangen \\ Simula Research Laboratory and University of Oslo}

\date{July 30, 2010} 
\maketitle


% lines beginning with # are comment lines

\begin{itemize}
 \item When writing a note, report, manual, etc., do you find it difficult
   to choose the typesetting format? That is, to choose between plain
   (email-like) text, Wiki, Word/OpenOffice, LaTeX, HTML,
   reStructuredText, Sphinx, XML, etc.  Would it be convenient to
   start with some very simple text-like format that easily converts
   to the formats listed above, and at some later stage eventually go
   with a particular format?

 \item Do you find it problematic that you have the same information
   scattered around in different documents in different typesetting
   formats? Would it be a good idea to write things once, in one
   place, and include it everywhere?

\end{itemize}
If any of these questions are of interest, you should keep on reading.

\subsection*{The Doconce Concept}

Doconce is two things:

\begin{enumerate}
 \item Doconce is a working strategy for documenting software in a single
    place and avoiding duplication of information. The slogan is:
    "Write once, include everywhere". This requires that what you
    write can be transformed to many different formats for a variety
    of documents (manuals, tutorials, books, doc strings, source code
    comments, etc.).

 \item Doconce is a simple and minimally tagged markup language that can
    be used for the above purpose. The Doconce format look
    like ordinary ASCII text (much like what you would use in an
    email), but the text can be transformed to numerous other formats,
    including HTML, Wiki, LaTeX, reStructuredText, Sphinx, XML,
    OpenOffice/Word, Epytext, PDF, XML - and even plain text (with
    tags removed for clearer reading).


\end{enumerate}

\subsection*{What Does Doconce Look Like?}

Doconce text looks like ordinary text, but there are some almost invisible
text constructions that allow you to control the formating. For example,

\begin{itemize}
  \item bullet lists arise from lines starting with an asterix,

  \item \emph{emphasized words} are surrounded by an asterix, 

  \item \textbf{words in boldface} are surrounded by underscores, 

  \item words from computer code are enclosed in back quotes and 
    then typeset verbatim,

  \item blocks of computer code can easily be included, also from source files,

  \item blocks of LaTeX mathematics can easily be included,

  \item there is support oforboth LaTeX and text-like inline mathematics,

  \item figures with captions, URLs with links, labels and references
    are supported,

  \item comments can be inserted throughout the text,

  \item a preprocessor (much like the C preprocessor) is integrated so
    other documents (files) can be included and large portions of text
    can be defined in or out of the text.

\end{itemize}
Here is an example of some simple text written in the Doconce format:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
===== A Subsection with Sample Text =====

Ordinary text looks like ordinary text, and the tags used for
_boldface_ words, *emphasized* words, and `computer` words look
natural in plain text.  Lists are typeset as you would do in an email,

  * item 1
  * item 2
  * item 3

Lists can also have automatically numbered items instead of bullets,

  o item 1
  o item 2
  o item 3

URLs with a link word are possible, as in http://folk.uio.no/hpl<hpl>.
Tables are also supperted, e.g.,

  |--------------------------------|
  |time  | velocity | acceleration |
  |--------------------------------|
  | 0.0  | 1.4186   | -5.01        |
  | 2.0  | 1.376512 | 11.919       |
  | 4.0  | 1.1E+1   | 14.717624    |
  |--------------------------------|
\end{Verbatim}
\noindent
The Doconce text above results in the following little document:

\subsection*{A Subsection with Sample Text}

Ordinary text looks like ordinary text, and the tags used for
\textbf{boldface} words, \emph{emphasized} words, and {\fontsize{10pt}{10pt}\verb!computer!} words look
natural in plain text.  Lists are typeset as you would do in an email,

\begin{itemize}
  \item item 1
  \item item 2
  \item item 3

\end{itemize}
Lists can also have numbered items instead of bullets, just use an
\begin{enumerate}
\item (for ordered) instead of the asterix:

\begin{enumerate}
 \item item 1
 \item item 2
 \item item 3

\end{enumerate}
\end{enumerate}
URLs with a link word are possible, as in \href{http://folk.uio.no/hpl}{hpl}.
Tables are also supperted, e.g.,


\begin{quote}\begin{tabular}{ccc}
\hline
\multicolumn{1}{c}{time} & \multicolumn{1}{c}{velocity} & \multicolumn{1}{c}{acceleration} \\
\hline
0.0          & 1.4186       & -5.01        \\
2.0          & 1.376512     & 11.919       \\
4.0          & 1.1E+1       & 14.717624    \\
\hline
\end{tabular}\end{quote}

\noindent

\subsection*{Mathematics and Computer Code}

Inline mathematics, such as $\nu = \sin(x)$,
allows the formula to be specified both as LaTeX and as plain text.
This results in a professional LaTeX typesetting, but in other formats
the text version normally looks better than raw LaTeX mathematics with
backslashes. An inline formula like $\nu = \sin(x)$ is
typeset as
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
$\nu = \sin(x)$|$v = sin(x)$
\end{Verbatim}
\noindent
The pipe symbol acts as a delimiter between LaTeX code and the plain text
version of the formula.

Blocks of mathematics are better typeset with raw LaTeX, inside
{\fontsize{10pt}{10pt}\verb!!bt!} and {\fontsize{10pt}{10pt}\verb!!et!} (begin tex / end tex) instructions. 
The result looks like this:

\begin{eqnarray}
{\partial u\over\partial t} &=& \nabla^2 u + f,\label{myeq1}\\
{\partial v\over\partial t} &=& \nabla\cdot(q(u)\nabla v) + g
\end{eqnarray}
Of course, such blocks only looks nice in LaTeX. The raw
LaTeX syntax appears in all other formats (but can still be useful
for those who can read LaTeX syntax).

You can have blocks of computer code, starting and ending with
{\fontsize{10pt}{10pt}\verb!!bc!} and {\fontsize{10pt}{10pt}\verb!!ec!} instructions, respectively. Such blocks look like
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
from math import sin, pi
def myfunc(x):
    return sin(pi*x)

import integrate
I = integrate.trapezoidal(myfunc, 0, pi, 100)
\end{Verbatim}
\noindent

One can also copy computer code directly from files, either the
complete file or specified parts.  Computer code is then never
duplicated in the documentation (important for the principle of
never copying anything!).

Another document can be included by writing {\fontsize{10pt}{10pt}\verb!#include "mynote.do.txt"!}
on a line starting with (another) hash sign.  Doconce documents have
extension {\fontsize{10pt}{10pt}\verb!do.txt!}. The {\fontsize{10pt}{10pt}\verb!do!} part stands for doconce, while the
trailing {\fontsize{10pt}{10pt}\verb!.txt!} denotes a text document so that editors gives you the
right writing enviroment for plain text.

\subsection*{Seeing More of What Doconce Is}

After the quick syntax tour above, we recommend to read the Doconce
source of the current tutorial and compare it with what you see in
a browser, a PDF document, in plain text, and so forth.
The Doconce source is found in the folder {\fontsize{10pt}{10pt}\verb!doc/tutorial.do.txt!} in the
source code tree of Doconce. The Doconce example documentation
displays both the source {\fontsize{10pt}{10pt}\verb!tutorial.do.txt!} and the result of many other
formats.

A more complete documentation of and motivation for Doconce appears
in the file {\fontsize{10pt}{10pt}\verb!lib/doconce/doc/doconce.do.txt!} in the Doconce
source code tree. The same documentation appears in the doc string of
the {\fontsize{10pt}{10pt}\verb!doconce!} module.

% Example on including another Doconce file:

\subsection*{From Doconce to Other Formats}

Transformation of a Doconce document to various other
formats applies the script {\fontsize{10pt}{10pt}\verb!doconce2format!}:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format format mydoc.do.txt
\end{Verbatim}
\noindent
The {\fontsize{10pt}{10pt}\verb!preprocess!} program is always used to preprocess the file first,
and options to {\fontsize{10pt}{10pt}\verb!preprocess!} can be added after the filename. For example,
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format LaTeX mydoc.do.txt -Dextra_sections
\end{Verbatim}
\noindent
The variable {\fontsize{10pt}{10pt}\verb!FORMAT!} is always defined as the current format when
running {\fontsize{10pt}{10pt}\verb!preprocess!}. That is, in the last example, {\fontsize{10pt}{10pt}\verb!FORMAT!} is
defined as {\fontsize{10pt}{10pt}\verb!LaTeX!}. Inside the Doconce document one can then perform
format specific actions through tests like {\fontsize{10pt}{10pt}\verb!#if FORMAT == "LaTeX"!}.
\paragraph{HTML.}

Making an HTML version of a Doconce file {\fontsize{10pt}{10pt}\verb!mydoc.do.txt!}
is performed by
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format HTML mydoc.do.txt
\end{Verbatim}
\noindent
The resulting file {\fontsize{10pt}{10pt}\verb!mydoc.html!} can be loaded into any web browser for viewing.
\paragraph{LaTeX.}

Making a LaTeX file {\fontsize{10pt}{10pt}\verb!mydoc.tex!} from {\fontsize{10pt}{10pt}\verb!mydoc.do.txt!} is done in two steps:
% Note: putting code blocks inside a list is not successful in many
% formats - the text may be messed up. A better choice is a paragraph
% environment, as used here.

\paragraph{Step 1.}
Filter the doconce text to a pre-LaTeX form {\fontsize{10pt}{10pt}\verb!mydoc.p.tex!} for
     {\fontsize{10pt}{10pt}\verb!ptex2tex!}:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format LaTeX mydoc.do.txt
\end{Verbatim}
\noindent
LaTeX-specific commands ("newcommands") in math formulas and similar
can be placed in a file {\fontsize{10pt}{10pt}\verb!newcommands.tex!}. If this file is present,
it is included in the LaTeX document so that your commands are
defined.

\paragraph{Step 2.}
Run {\fontsize{10pt}{10pt}\verb!ptex2tex!} (if you have it) to make a standard LaTeX file,
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> ptex2tex mydoc
\end{Verbatim}
\noindent
     or just perform a plain copy,
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> cp mydoc.p.tex mydoc.tex
\end{Verbatim}
\noindent
The {\fontsize{10pt}{10pt}\verb!ptex2tex!} tool makes it possible to easily switch between many
different fancy formattings of computer or verbatim code in LaTeX
documents.
Finally, compile {\fontsize{10pt}{10pt}\verb!mydoc.tex!} the usual way and create the PDF file.
\paragraph{Plain ASCII Text.}

We can go from Doconce "back to" plain untagged text suitable for viewing
in terminal windows, inclusion in email text, or for insertion in
computer source code:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format plain mydoc.do.txt  # results in mydoc.txt
\end{Verbatim}
\noindent
\paragraph{reStructuredText.}

Going from Doconce to reStructuredText gives a lot of possibilities to
go to other formats. First we filter the Doconce text to a
reStructuredText file {\fontsize{10pt}{10pt}\verb!mydoc.rst!}:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format rst mydoc.do.txt
\end{Verbatim}
\noindent
We may now produce various other formats:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> rst2html.py  mydoc.rst > mydoc.html # HTML
Unix/DOS> rst2latex.py mydoc.rst > mydoc.tex  # LaTeX
Unix/DOS> rst2xml.py   mydoc.rst > mydoc.xml  # XML
Unix/DOS> rst2odt.py   mydoc.rst > mydoc.odt  # OpenOffice
\end{Verbatim}
\noindent
The OpenOffice file {\fontsize{10pt}{10pt}\verb!mydoc.odt!} can be loaded into OpenOffice and
saved in, among other things, the RTF format or the Microsoft Word format.
That is, one can easily go from Doconce to Microsoft Word.
\paragraph{Sphinx.}

Sphinx documents can be created from a Doconce source in a few steps.

\paragraph{Step 1.}
Translate Doconce into the Sphinx dialect of
the reStructuredText format:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> doconce2format sphinx mydoc.do.txt
\end{Verbatim}
\noindent

\paragraph{Step 2.}
Create a Sphinx root directory with a {\fontsize{10pt}{10pt}\verb!conf.py!} file, 
either manually or by using the interactive {\fontsize{10pt}{10pt}\verb!sphinx-quickstart!}
program. Here is a scripted version of the steps with the latter:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
mkdir sphinx-rootdir
sphinx-quickstart <<EOF
sphinx-rootdir
n
_
Name of My Sphinx Document
Author
version
version
.rst
index
y
n
n
n
n
y
n
n
y
y
EOF
\end{Verbatim}
\noindent

\paragraph{Step 3.}
Move the {\fontsize{10pt}{10pt}\verb!tutorial.rst!} file to the Sphinx root directory:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> mv mydoc.rst sphinx-rootdir
\end{Verbatim}
\noindent

\paragraph{Step 4.}
Edit the generated {\fontsize{10pt}{10pt}\verb!index.rst!} file so that {\fontsize{10pt}{10pt}\verb!mydoc.rst!}
is included, i.e., add {\fontsize{10pt}{10pt}\verb!mydoc!} to the {\fontsize{10pt}{10pt}\verb!toctree!} section so that it becomes
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
.. toctree::
   :maxdepth: 2

   mydoc
\end{Verbatim}
\noindent
(The spaces before {\fontsize{10pt}{10pt}\verb!mydoc!} are important!)

\paragraph{Step 5.}
Generate, for instance, an HTML version of the Sphinx source:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
make clean   # remove old versions
make html
\end{Verbatim}
\noindent
Many other formats are also possible.

\paragraph{Step 6.}
View the result:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
Unix/DOS> firefox _build/html/index.html
\end{Verbatim}
\noindent
\paragraph{Demos.}

The current text is generated from a Doconce format stored in the file
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
tutorial/tutorial.do.txt
\end{Verbatim}
\noindent
The file {\fontsize{10pt}{10pt}\verb!make.sh!} in the {\fontsize{10pt}{10pt}\verb!tutorial!} directory of the
Doconce source code contains a demo of how to produce a variety of
formats.  The source of this tutorial, {\fontsize{10pt}{10pt}\verb!tutorial.do.txt!} is the
starting point.  Running {\fontsize{10pt}{10pt}\verb!make.sh!} and studying the various generated
files and comparing them with the original {\fontsize{10pt}{10pt}\verb!doconce.do.txt!} file,
gives a quick introduction to how Doconce is used in a real case.

There is another demo in the {\fontsize{10pt}{10pt}\verb!lib/doconce/doc!} directory which
translates the more comprehensive documentation, {\fontsize{10pt}{10pt}\verb!doconce.do.txt!}, to
various formats.  For example, to go from the LaTeX format to PDF, see
{\fontsize{10pt}{10pt}\verb!latex.sh!}.

\subsection*{The Doconce Documentation Strategy for User Manuals}

Doconce was particularly made for writing tutorials or user manuals
associated with computer codes. The text is written in Doconce format
in separate files. LaTeX, HTML, XML, and other versions of the text
is easily produced by the {\fontsize{10pt}{10pt}\verb!doconce2format!} script and standard tools.
A plain text version is often wanted for the computer source code,
this is easy to make, and then one can use
{\fontsize{10pt}{10pt}\verb!#include!} statements in the computer source code to automatically
get the manual or tutorial text in comments or doc strings.
Below is a worked example.

Consider an example involving a Python module in a {\fontsize{10pt}{10pt}\verb!basename.p.py!} file.
The {\fontsize{10pt}{10pt}\verb!.p.py!} extension identifies this as a file that has to be
preprocessed) by the {\fontsize{10pt}{10pt}\verb!preprocess!} program. 
In a doc string in {\fontsize{10pt}{10pt}\verb!basename.p.py!} we do a preprocessor include
in a comment line, say
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
#    #include "docstrings/doc1.dst.txt
\end{Verbatim}
\noindent
% 
% Note: we insert an error right above as the right quote is missing.
% Then preprocess skips the statement, otherwise it gives an error
% message about a missing file docstrings/doc1.dst.txt (which we don't
% have, it's just a sample file name). Also note that comment lines
% must not come before a code block for the rst/st/epytext formats to work.
% 
The file {\fontsize{10pt}{10pt}\verb!docstrings/doc1.dst.txt!} is a file filtered to a specific format
(typically plain text, reStructedText, or Epytext) from an original
"singleton" documentation file named {\fontsize{10pt}{10pt}\verb!docstrings/doc1.do.txt!}. The {\fontsize{10pt}{10pt}\verb!.dst.txt!}
is the extension of a file filtered ready for being included in a doc
string ({\fontsize{10pt}{10pt}\verb!d!} for doc, {\fontsize{10pt}{10pt}\verb!st!} for string).

For making an Epydoc manual, the {\fontsize{10pt}{10pt}\verb!docstrings/doc1.do.txt!} file is
filtered to {\fontsize{10pt}{10pt}\verb!docstrings/doc1.epytext!} and renamed to
{\fontsize{10pt}{10pt}\verb!docstrings/doc1.dst.txt!}.  Then we run the preprocessor on the
{\fontsize{10pt}{10pt}\verb!basename.p.py!} file and create a real Python file
{\fontsize{10pt}{10pt}\verb!basename.py!}. Finally, we run Epydoc on this file. Alternatively, and
nowadays preferably, we use Sphinx for API documentation and then the
Doconce {\fontsize{10pt}{10pt}\verb!docstrings/doc1.do.txt!} file is filtered to
{\fontsize{10pt}{10pt}\verb!docstrings/doc1.rst!} and renamed to {\fontsize{10pt}{10pt}\verb!docstrings/doc1.dst.txt!}. A
Sphinx directory must have been made with the right {\fontsize{10pt}{10pt}\verb!index.rst!} and
{\fontsize{10pt}{10pt}\verb!conf.py!} files. Going to this directory and typing {\fontsize{10pt}{10pt}\verb!make html!} makes
the HTML version of the Sphinx API documentation.

The next step is to produce the final pure Python source code. For
this purpose we filter {\fontsize{10pt}{10pt}\verb!docstrings/doc1.do.txt!} to plain text format
({\fontsize{10pt}{10pt}\verb!docstrings/doc1.txt!}) and rename to {\fontsize{10pt}{10pt}\verb!docstrings/doc1.dst.txt!}. The
preprocessor transforms the {\fontsize{10pt}{10pt}\verb!basename.p.py!} file to a standard Python
file {\fontsize{10pt}{10pt}\verb!basename.py!}. The doc strings are now in plain text and well
suited for Pydoc or reading by humans. All these steps are automated
by the {\fontsize{10pt}{10pt}\verb!insertdocstr.py!} script.  Here are the corresponding Unix
commands:
\begin{Verbatim}[fontsize=\fontsize{9pt}{9pt},tabsize=8,baselinestretch=0.85,
fontfamily=tt,xleftmargin=7mm]
# make Epydoc API manual of basename module:
cd docstrings
doconce2format epytext doc1.do.txt
mv doc1.epytext doc1.dst.txt
cd ..
preprocess basename.p.py > basename.py
epydoc basename

# make Sphinx API manual of basename module:
cd doc
doconce2format sphinx doc1.do.txt
mv doc1.rst doc1.dst.txt
cd ..
preprocess basename.p.py > basename.py
cd docstrings/sphinx-rootdir  # sphinx directory for API source
make clean
make html
cd ../..

# make ordinary Python module files with doc strings:
cd docstrings
doconce2format plain doc1.do.txt
mv doc1.txt doc1.dst.txt
cd ..
preprocess basename.p.py > basename.py

# can automate inserting doc strings in all .p.py files:
insertdocstr.py plain .
# (runs through all .do.txt files and filters them to plain format and
# renames to .dst.txt extension, then the script runs through all 
# .p.py files and runs the preprocessor, which includes the .dst.txt
# files)
\end{Verbatim}
\noindent


\section*{Warning/Disclaimer}

Doconce can be viewed is a unified interface to a variety of
typesetting formats.  This interface is minimal in the sense that a
lot of typesetting features are not supported, for example, footnotes
and bibliography. For many documents the simple Doconce format is
sufficient, while in other cases you need more sophisticated
formats. Then you can just filter the Doconce text to a more
approprite format and continue working in this format only.  For
example, reStructuredText is a good alternative: it is more tagged
than Doconce and cannot be filtered to plain, untagged text, or wiki,
and the LaTeX output is not at all as clean, but it also has a lot
more typesetting and tagging features than Doconce.\end{document}